Keep your Laziness in Check

### 2018-03-05 paper/main.tex
[paper] Intro/Sec 2 nits

Lazy evaluation gives tremendous  to functional programmers
~\cite{WhyFPMatters} and is actively used in various practical
applications. For instance, the big-data framework Spark uses laziness
to drive computations on demand\cn, Apple's programming language Swift
leverages it to initialize expensive resources as needed\cn, and many
functional reactive programming web frameworks crucially rely on
it\cn \leo{This point might need more exposition about frp}. Haskell
is the most popular programming language that allows programmers to
enjoy the power of lazy evaluation by default. However, that power
does not come without a price.
\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
bugs or something? Work that in?}
\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
get right, but let's check with him on whether we can quote him on that.}

\caleb{Again, power and responsibility are not completely clear, but I do like the narrative}
The power of laziness puts great responsibility on programmers'
shoulders. An incorrect implementation can either behave functionally
the same but with inferior \caleb{Inferior than what?} performance, or it can completely break
down!\caleb{unclear. loops?} Reasoning about lazy programs is also notoriously hard --- the
programmer needs to have a mental model of the order of evaluation of
the code, and this order is not always obvious from the syntax of the
code.

How, then, can we test and catch these laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following small example on the
function \lk{take}, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict (nf n) (if n > length xs then E d else endThunk d)
  where
    endThunk :: PosDemand [a] -> Demand [a]
    endThunk = \case
      Nil'        -> T
      dx :% T     -> E (dx :% T)
      dx :% E dxs -> E (dx :% endThunk dxs)
\end{inlinecode}
%
\hzh{TODO: describe what this spec means in English}

In the next section we work through two more examples, functional
queues and functional maps, demonstrating both the difficulties of
programming with laziness and how our approach can be used to
alleviate them.
%
In the rest of the paper, we present our technical contributions:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior
of functions that is {\em precise} and {\em compositional}. Such
specifications are expressive enough to capture exactly how much
evaluation occurs within a data structure
%\leo{Also add
%drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm
%not sure if we should bring in partiality (in the sense of
%divergence/nontermination) yet since that seems like a rather
%technical detail of the spec}
\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}
\item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\caleb{I don't know
the meaning of non-trivial and almost-all}
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

@@

@@

Turning our intuition above into a Haskell function, we complete the
specification of \lk{rotate} in \Cref{fig:rotate-spec}
\begin{figure}[H]
\centering
\hzh{TODO}
\begin{inlinecode}
TODO: rewrite this spec so it looks like the invariant in Okasaki's paper.
\end{inlinecode}
\caption{Specification for \lk{rotate}}
\label{fig:rotate-spec}
\end{figure}

\lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example. \hzh{should we show what a
counter example looks like?}\leo{Definitely!}

@@

For the second example, we turn to a different functional data structure where
both a strict and a lazy variant can be situationally useful: maps. There are
two reasonable laziness behaviors that implement a map interface; we could
implement a completely strict map, or we could lazily evaluate the values stored
as needed.
\footnote{A third option could be to have lazy keys, but they would
necessarily be forced at every insertion or lookup, with no gain.}
Haskell's standard library provides these two variants in the form of
\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
to {\em specify} the difference between the two versions! At best, one
could hope for a hand-crafted suite of unit tests containing divergent
terms to observe a difference in termination.
\leo{One thing we could do is ignore the knapsack and expand on the above.}

>>>>

Lazy evaluation gives tremendous power to functional programmers
~\cite{WhyFPMatters} and is actively used in various practical
applications. For instance, the big-data framework Spark uses laziness
to drive computations on demand\cn, Apple's programming language Swift
leverages it to initialize expensive resources as needed\cn, and many
functional reactive programming web frameworks rely on
it crucially\cn \leo{This point might need more exposition about frp}. Haskell
is the most popular programming language that allows programmers to
enjoy the power of lazy evaluation by default. However, it
does not come without a price.
\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
bugs or something? Work that in?}
\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
get right, but let's check with him on whether we can quote him on that.}

\caleb{Again, power and responsibility are not completely clear, but I do like the narrative}
The power of laziness puts great responsibility on programmers'
shoulders. An implementation that uses laziness incorrectly can either
behave functionally the same but with inferior \caleb{Inferior than
what?} performance---compared to a strict counterpart, or it can
completely break down!\caleb{unclear. loops?} Reasoning about lazy
programs is also notoriously hard --- the programmer needs to have a
mental model of the order of evaluation of the code, and this order is
not always obvious from the syntax of the code.

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict (nf n) (if n > length xs then E d else endThunk d)
  where
    endThunk :: PosDemand [a] -> Demand [a]
    endThunk = \case
      Nil'        -> T
      dx :% T     -> E (dx :% T)
      dx :% E dxs -> E (dx :% endThunk dxs)
\end{inlinecode}
%
\hzh{TODO: describe what this spec means in English}

In the next section we work through two more examples, functional
queues and functional maps, demonstrating both the difficulties of
programming with laziness and how our approach can be used to
alleviate them.
%
In the rest of the paper, we present our technical contributions:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior
of functions that is {\em precise} and {\em compositional}. Such
specifications are expressive enough to capture exactly how much
evaluation occurs within a data structure
%\leo{Also add
%drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm
%not sure if we should bring in partiality (in the sense of
%divergence/nontermination) yet since that seems like a rather
%technical detail of the spec}
\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}\leo{Should we mention QuickCheck here?}
\item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\leo{I guess ``packed'' maps do not qualified as algebraic?}
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

@@

\subsection{Purely functional queues}
\label{sec:queues}

@@

Turning our intuition above into a Haskell function, we complete the
specification of \lk{rotate} in \Cref{fig:rotate-spec}.
%
\begin{figure}[H]
\centering
\hzh{TODO}
\begin{inlinecode}
TODO: rewrite this spec so it looks like the invariant in Okasaki's paper.
\end{inlinecode}
\caption{Specification for \lk{rotate}}
\label{fig:rotate-spec}
\end{figure}
%
The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example. \hzh{should we show what a
counter example looks like?}\leo{Definitely!}

\subsection{Key- vs value-strict maps}

@@

For the second example, we turn to a different functional data structure where
both a strict and a lazy variant can be situationally useful: maps. There are
two reasonable laziness behaviors that implement a map interface; we could
implement a completely strict map, or we could lazily evaluate the values stored
as needed.
\footnote{A third option could be to have lazy keys, but they would
necessarily be forced at every insertion or lookup, with no gain.}
Haskell's standard library provides these two variants in the form of
\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
to {\em specify} the difference between the two versions! At best, one
could hope for a hand-crafted suite of unit tests containing divergent
terms to observe a difference in termination.
\leo{One thing we could do is ignore the knapsack and expand on the above.}
\leo{Here we need a similar discussion as with lists}
